Nathan Greenslit Data Analysis Project

### 2023-04-05 code/analysis_code/stats/2_cca.qmd
Project edits

```{r, include = FALSE}
library(tidyverse)
library(here)
```

@@

>>>>

```{r, include = FALSE}
library(tidyverse)
library(here)
library(ggpubr)
```

@@

#### Save Figures

```{r}
jpeg(file = here("results", "plots", "cca", "cca_bo.jpg"))
bo_ccf<- bo %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19'))) 
  ccf(bo_ccf$t7sum, bo_ccf$copies_mL) #CCA between dust and copies_mL
  title("Blind Oso DustxCopies CCA", line =1)
dev.off()

jpeg(file = here("results", "plots", "cca", "cca_c2.jpg"))
c2_ccf<- c2 %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
  ccf(c2_ccf$t7sum, c2_ccf$copies_mL)
  title("Canals DustxCopies CCA", line=1)
dev.off()

jpeg(file = here("results", "plots", "cca", "cca_rd.jpg"))
rd_ccf<- rd %>%
  filter(between(date, as.Date('2022-07-07'), as.Date('2022-07-19')))
  ccf(rd_ccf$t7sum, rd_ccf$copies_mL) 
  title("Gulf DustxCopies CCA", line =1)
dev.off()
```

### 2023-04-05 code/analysis_code/stats/3_uni_variate.qmd
Project edits

```{r}
bo2 %>% ggplot(aes(x = lag, y = copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (BO+1)",
       x = "log(dust)")

c22 %>% ggplot(aes(x = lag, y = copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (C2+2)",
       x = "log(dust)")

rd2 %>% ggplot(aes(lag, copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (RD+2)",
       x = "log(dust)")
```

>>>>

```{r}
fig1<- bo2 %>% ggplot(aes(x = lag, y = copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (BO+1)",
       x = "log(dust)")

fig2<- c22 %>% ggplot(aes(x = lag, y = copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (C2+2)",
       x = "log(dust)")

fig3<- rd2 %>% ggplot(aes(lag, copies_mL)) +
  geom_point() +
  stat_smooth(method = "lm")+
  stat_poly_eq(aes(x = lag, y = copies_mL),
               rr.digits = 4) +
  labs(title = "LM DustxCopies (RD+2)",
       x = "log(dust)")
```

#### Stack and Save

```{r}
png(file = here("results", "plots", "stats", "lm.png"))
ggarrange(fig1, fig2, fig3 + font("x.text", size = 10), ncol = 1, nrow = 3)
dev.off()
```

### 2023-04-05 products/manuscript/Manuscript.qmd
Project edits

Each year, plumes of
Saharan dust travel across the Atlantic to be deposited in the surface waters of the Caribbean and Gulf of Mexico. Dust aerosols serve as a significant source of nutrients that can elicit a fertilization effect on marine coastal waters, leading to rapid and potentially harmful blooms (e.g. red tides). Previous work has characterized the response of the marine bacteria *Vibrio* to dust input under the oligotrophic settings of the Florida Keys, but less is (1) quantify *Vibrio* population dynamics and composition, and (2) characterize
potential microbial blooms and community shifts in response to dust input in coastal sites with higher ambient nutrient levels. Monthly and daily sampling influences on microbialpopulations and to capture before, during, and after a Saharan dust event. Culturable *Vibrio* were enumerated as CFU ml^-1^ via spread plating on selective media. qPCRwill be used to estimate total counts of *Vibrio* and of pathogenic *Vibrio* species (*V. cholerae, V. parahaemolyticus, V. vulnificus,* and *V. alginolyticus*) while *Vibrio* composition will be determined using hsp60 gene or 16S rRNA amplicon sequencing. Microbial community composition will be determined using 16S rRNA and 18S rRNA gene amplicon sequencing to examine shifts in prokaryotic and eukaryotic composition over time. This study will provide an increasedunderstanding of the conditions that can elicit potentially harmful blooms, highlighting the need for further research to understand effects of dust

@@

Each year, plumes of Saharan dust travel across the Atlantic via easterly trade winds. These plumes are then deposited in the surface waters of the mid-Atlantic, Caribbean, and Gulf of Mexico. These dust aerosols harbor a wide range of fungi, bacteria, virus-like-particles, minerals, and nutrients (NO3, Fe, PO4). Addition of these nutrients to otherwise oligotrophic settings can result in large and rapid blooms of potentially harmful microbes, presenting a danger to both marine and human health.

Bacteria belonging to the genus Vibrio are marine opportunistic heterotrophs that are ubiquitous in nature and are among the first to respond to the influx of nutrients. Species of Vibrio are known to cause a variety of diseases in humans ranging from cholera (V. cholerae), to ear and wound infections (V. alginolyticus), and in some extreme cases necrotizing fasciitis and septicemia (V. vulnificus). Vibrio have also been shown to impact marine health, causing coral bleaching (V. coralliiyticus, V. shiloi, V. alginolyticus), and shellfish and fish mortality (V. harveyi, V. parahaemolyticus, V. vulnificus). Vibrio population dynamics are primarily driven by temperature and salinity, with prime conditions between 25-30°C and salinities ranging from 20-35. While there is evidence of these blooms occurring in low-nutrient settings like the Florida Keys, less is known regarding microbial response to dust input in areas with *higher baseline nutrients*.

@@

## Data acquisition

Data was collected at 3 sites in Corpus Christi, TX before, during, and after a Saharan dust event (starting on July 7th and ending on July 19th, 2022). Sites represent a gradient in background nutrient levels (Blind Oso and Canals are high, the Gulf is low). Water samples were collected for inorganic nutrients, dissolved and particulate organic matter, and microbial analysis. Vertical profiles of salinity, temperature, pH, and dissolved oxygen were obtained using a YSI ProPlus sonde.

Variables will include `Temperature (°C)`, `Salinity`, `Nitrate`, `Phosphate`, `Copies/mL` produced using quantitative PCR, and `Dust Concentration (AOT)` from the naval research laboratory satellite data.

@@

This project works with three primary data sets. (1) Enumeration of Vibrio bacteria as copies/mL from quantitative PCR (copies_master.csv), (2) Dust concentrations as aerosol optical thickness (AOT) derived from the Naval Research Lab (nrl_conc.csv), and (3) Temperature and Salinity across the daily time series (ysi_2.csv). Below is a brief summary of how each data set was cleaned. More details can be found in the supplementary files, containing the code and comments describing what each line does. Please refer to the `project_README.md` file located in the NathanGreenslit-MADA-Project folder for details on the repository contents and instructions on reproducibility

### Cleaning of Total Vibrio Enumerations Data set

@@

### Cleaning of Dust Concentration (AOT) Data set

@@

### Cleaning of Temperature and Salinity Data set

@@

### Dust

@@

### Daily Vibrio

@@

### Temperature and Salinity

@@

@@

<!--# Add distribution statement -->

@@

![](images/00003f.png){width="475"}

![](images/000043.png){width="473"}

![](images/000047.png){width="473"}

@@

<!--# Need to update these plots -->

![](images/00000f.png){width="382"}

![](images/00000b.png){width="381"}

![](images/000013.png){width="381"}

>>>>

Each year, plumes of Saharan dust travel across the Atlantic to be deposited in the surface waters of the Caribbean and Gulf of Mexico. Dust aerosols serve as a significant source of nutrients that can elicit a fertilization effect on marine coastal waters, leading to rapid and potentially harmful blooms (e.g. red tides). Previous work has characterized the response of the marine bacteria *Vibrio* to dust input under the oligotrophic settings of the Florida Keys, but less is known regarding settings with higher baseline nutrient levels. This project aims to quantify *Vibrio* population dynamics in response to dust input in coastal sites with higher ambient nutrient levels. Daily sampling took place in Corpus Christi, TX to capture before, during, and after a Saharan dust event and quantitative PCR (qPCR) was used to estimate total counts of *Vibrio.* This study will provide an increased understanding of the conditions that can elicit potentially harmful blooms, highlighting the need for further research to understand the effects of dust deposition in non-oligotrophic waters.

@@

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year ([d'Almeida, 1986](https://journals.ametsoc.org/view/journals/apme/25/7/1520-0450_1986_025_0903_amfsdt_2_0_co_2.xml)). Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico (Fig. 1). These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria ([Kellogg et al., 2004](https://link.springer.com/content/pdf/10.1023/B:AERO.0000032947.88335.bb.pdf)), fungi ([Ramírez Camejo et al., 2022](https://link.springer.com/article/10.1007/s10453-022-09752-9)), virus-like particles ([Griffin et al., 2001](https://link.springer.com/article/10.1023/A:1011868218901)), and minerals ([Formenti et al., 2003](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2002JD002648)). These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron ([Graham & Duce, 1967](https://www.sciencedirect.com/science/article/pii/0004698182901986), [Mills et al., 2004](https://pubmed.ncbi.nlm.nih.gov/15152251/), [Savoie et al., 1989](https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/JD094iD04p05069?casa_token=PMRaet9QSYsAAAAA:nb8yBorxdKaEWwGfI5aNILJ0D4h6drNu2YY1y1y8ExbQ03z59EUTQTOHoxinzsxRtR4jzsfSA73_Hig)). This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups ([Lenes et al., 2008](https://www.sciencedirect.com/science/article/pii/S0278434308000575), [Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113), [Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)).

Previous work conducted in the West Florida Shelf has shown that increased iron availability can stimulate the growth of the nitrogen-fixing cyanobacterium *Trichodesmium*. The high amount of nitrogen produced via fixation can be adequate enough to stimulate toxic red tides, caused by the dinoflagellate *Karenia brevis* ([Lenes et al., 2008](https://www.sciencedirect.com/science/article/pii/S0278434308000575)).

While it is documented that these dust events can stimulate growth of harmful phytoplankton, emerging evidence shows that heterotrophic bacteria are responding to the episodic influx of limiting resources and substrate, resulting in blooms of bacteria that are associated with disease, which we refer to as harmful bacterial blooms. Among those that respond are bacterium belonging to the genus *Vibrio*, a group of ubiquitous marine opportunistic heterotrophs. While considered conditionally rare (comprising of \<1% of the microbial community), under the proper conditions, this genus can rapidly bloom to make up a large percentage of the community over a short period of time ([Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113), [Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). *Vibrio* species are found in brackish to marine waters with salinities ranging from 0-35 and will grow in waters with temperatures ranging from 5°C-40°C ([Sampaio et al., 2022](https://www.mdpi.com/1424-2818/14/2/97)). Each year, plumes of Saharan dust travel across the Atlantic via easterly trade winds. These plumes are then deposited in the surface waters of the mid-Atlantic, Caribbean, and Gulf of Mexico. These dust aerosols harbor a wide range of fungi, bacteria, virus-like-particles, minerals, and nutrients (NO3, Fe, PO4). Addition of these nutrients to otherwise oligotrophic settings can result in large and rapid blooms of potentially harmful microbes, presenting a danger to both marine and human health.

*Vibrio* species are excellent opportunists, having multiple advantages that allow them to quickly respond to any newly introduced substrate or nutrients. These advantages include multiple copies of rRNA genes, a rapid doubling time, efficient chemotactic motility, large genomic repertoire (consisting of two circular chromosomes), and multiple Fe-siderophore complexes that allow for rapid uptake of iron into the cell ([Jensen et al., 2009](https://academic.oup.com/femsle/article/294/2/207/512983), [Ringgaard et al., 2018](https://journals.asm.org/doi/10.1128/JB.00793-17), [Eagon, 1962](https://journals.asm.org/doi/abs/10.1128/jb.83.4.736-737.1962?casa_token=WdtRNnQWyhkAAAAA:Zre66xvwgEu3wXQykenhnSoy3OA08h1u-4Zl31k9cxxxfV6jq46dzL18Y02OmwPHdY7p7tgH0wVN-4I), [Okada et al., 2005](https://journals.asm.org/doi/10.1128/JB.187.2.752-757.2005), [Payne et al., 2015](https://journals.asm.org/doi/full/10.1128/MMBR.00046-15)).

Within the *Vibrionaceae* family, there are at least 12 species that are commonly known as human pathogens. Illness can be induced through either a foodborne route via the consumption of raw or uncooked seafood and contaminated drinking water, or through nonfoodborne routes such as wound exposure while swimming. The most common illnesses caused by *Vibrio* include self-limiting diarrhea and cholera (*V. cholerae*), shellfish-induced gastroenteritis (*V. parahaemolyticus, V. vulnificus*), extreme cases of necrotizing fasciitis and septicemia (*V. vulnificus*), and wound and ear infections  (*V. alginolyticus*, *V. vulnificus*) ([Baker-Austin et al., 2018](https://www.nature.com/articles/s41572-018-0005-8)). Some species also pose a threat to marine health, inducing mortality in oyster and clam larvae (*V. tubiashii, V. coralliilyticus*), bacterial bleaching and rapid tissue loss in corals (*V. mediterranei, V. coralliilyticus*) ([Richards et al., 2015](https://pubmed.ncbi.nlm.nih.gov/25344234/), [Rosenberg & Falkovitz, 2004](https://www.annualreviews.org/doi/10.1146/annurev.micro.58.030603.123610), [Ben-Haim et al., 2003](https://pubmed.ncbi.nlm.nih.gov/12656189/)), and mortality in important aquaculture species such as penaeid shrimp (*V. parahaemolyticus*) and sea breams (*V. harveyi)* ([Tran et al., 2013](https://www.int-res.com/abstracts/dao/v105/n1/p45-55/), [Haldar et al., 2010](https://www.sciencedirect.com/science/article/pii/S0944501309001177)), among others.

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours (Fig. 2). *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community ([Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113)). A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition ([Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups ([Borchardt et al., 2019](https://aslopubs.onlinelibrary.wiley.com/doi/full/10.1002/lno.11291)). To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

This study aims to (1) quantify *Vibrio* population dynamics and composition, and (2) characterize potential microbial blooms and community shifts in response to dust input in coastal sites with higher ambient nutrient levels. The findings from this study will provide an increased understanding of the conditions that can elicit potentially harmful blooms (i.e., *Vibrio*), highlighting the need for further research to understand the effects of dust deposition in non-oligotrophic waters. We predict that (1) *Vibrio* will be among the first to respond to the influx of dust constituents and (2) that the fertilization effects of dust deposition (and possible successional changes) will be dampened at sites with higher baseline nutrient levels, as dust derived nutrients become less critical for growth.

With a heightened risk of *Vibrio* exposure in coastal waters, it is vital to understand microbial dynamics in nearshore environments. If dust is proven to elicit a microbial response, these events can be used to serve as an "early warning" of increased exposure risk to better protect public health.

@@

## Description of Study Area

Sampling took place at three locations in Corpus Christi, TX: Blind Oso Bay, a residential canal system on Padre Island, and the Gulf. Sites were chosen based on differing baseline nutrient levels. Blind Oso Bay is a shallow tributary that is popular amongst wadefisherman and kayakers. The Bay is often impacted by fluctuating salinities and high levels of nutrients and chlorophyl due to its proximity to a wastewater treatment plant ([Wetz, 2014](https://www.cbbep.org/manager/wp-content/uploads/1312-Oso-Bay-Nutrient-Assessment-Final_Report-6-24-14.pdf)). Additionally, the Bay has persistent issues with high levels of fecal indicator bacteria, placing it on the U.S. impaired waters list ([Nicolau & Hill, 2014](https://www.tceq.texas.gov/downloads/water-quality/tmdl/oso-bay-oyster-harvesting-assessment-103/103-oso-bay-monitoring-report-2013.pdf), [Texas Commission on Environmental Quality, 2022](https://www.tceq.texas.gov/downloads/water-quality/assessment/integrated-report-2022/2022-303d.pdf)). The residential canal system is also impacted by high nutrients and chlorophyl from storm water runoff, and salinity is primarily driven by precipitation. The Gulf site is characterized by constant salinities (\~36) and low nutrient levels.

## Sample Collection and Processing

#### Collection

A high resolution (daily) time series was conducted starting on July 7th and ending on July 19th, 2022 capturing before, during, and after a Saharan dust event. <!--# Add protocol -->

Vertical profiles of salinity, temperature, pH, and dissolved oxygen were obtained using a YSI ProPlus sonde. Samples were collected for chlorophyl *a* analysis, inorganic nutrients, dissolved and particulate organic matter, and microbial analysis.

#### Vibrio Enumeration

Total *Vibrio* were estimated using genus-specific quantitative polymerase chain reaction (qPCR) ([Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). <!--# Add specific primers, conditions, etc  -->

@@

This project works with three primary data sets. (1) Enumeration of Vibrio bacteria as copies/mL from quantitative PCR (copies_master.csv), (2) Dust concentrations as aerosol optical thickness (AOT) derived from the Naval Research Lab (nrl_conc.csv), and (3) Temperature and Salinity across the daily time series (ysi_2.csv). Below is a brief summary of how each data set was cleaned. More details can be found in the supplementary files, containing the code and comments describing what each line does. Please refer to the `project_README.md` file located in the NathanGreenslit-MADA-Project folder for details on the repository contents and instructions on reproducibility\

#### Cleaning of Total Vibrio Enumeration dataset

@@

#### Cleaning of Dust Concentration (AOT) Data set

@@

#### Cleaning of Environment data set

@@

#### Dust

@@

#### Daily Vibrio

@@

#### Temperature & Salinity

@@

#### Other Environmental Parameters

@@

Shapiro-Wilk's tests were conducted to assess distribution of the data. Any data that did not have a normal distribution was log-transformed and the test was run again.

@@

![](images/cca_bo.jpg){width="203"}

![](images/cca_c2.jpg){width="203"}

![](images/cca_rd.jpg){width="203"}

@@

![](images/lm.png){width="329"}
